pirouette: Determining the error BEAST2 makes in inferring a phylogeny

### 2019-03-13 article.tex
Use '~\ref' instead of ' \ref', as the tilde marks a non-breaking space. Thanks @Giappo

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to measure the inference error BEAST2
makes for a given reconstructed phylogeny, 
simulated under the mechanisms of a (possibly new) diversification model, 
which we will refer to as the 'generative prior'. 
Such a study will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the error made by BEAST2 cannot be neglected. 
Although a well-conducted theoretical study uses many replicates, 
this article only shows examples with one replicate, for the sake of clarity.

\verb;pirouette; is very flexible and allows the user 
to specify a big variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in Tables~\ref{tab:options} and \ref{tab:definitions}.

@@

An overview of \verb;pirouette;'s main functions is shown in 
table \ref{tab:functions}. 
We will exploit these functions to answer our research questions in the next sections.

@@

\begin{lstlisting}[
    language=R, floatplacement=ht, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in figure \ref{fig:yule_tree}
    }
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_yule.png}
  \caption{The Yule tree, as created by listing \ref{lst:create_yule_tree}.}
  \label{fig:yule_tree}
  \giovanni{Remove the text in top-left corner.}
  \giovanni{I'd use a sexier topology, purely for esthetical reasons.}
\end{figure}

@@

The resulting figure is shown in figure \ref{fig:example_1}.

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_unknown.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing \ref{lst:unknown_phylogeny}.
  }
\end{figure}

The first step in \verb;pirouette; is to simulate a DNA alignment from the given phylogeny. 
We will re-use the alignment parameters of the previous example 
as shown in listing \ref{lst:create_alignment}.

@@

The resulting figure is shown in figure \ref{fig:example_2}.

@@

Most of the settings for this example research question are
the same as in the previous section, which are
the phylogeny from an unknown speciation model (listing \ref{lst:unknown_phylogeny}), 
the alignment parameters (listing \ref{lst:create_alignment}) 
and the list of experiments (listing \ref{lst:create_generative_experiment}).

@@

The resulting figure is shown in figure \ref{fig:example_3}

@@

Here, same as example 2 and 3, we start with the phylogeny 
from an unknown speciation model (listing \ref{lst:unknown_phylogeny}), 
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing \ref{lst:create_alignment}.

In the second step we specify we want all 
40 inference models to compete, by 
calling \verb;create_all_experiments;,
as done in listing \ref{lst:create_all_experiments}.

@@

To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using the
default created twinning parameters, as shown in 
listing \ref{lst:create_twinning_params}.

@@

The resulting figure is shown in figure \ref{fig:example_4}.

@@

In this example, we use a tree generated by the Yule process.
Instead of using the common Yule tree as in example 1, we'll use
a Yule tree of low (actually: zero) likelihood: 
the one as used in examples 2-4.
by using the code shown in listing \ref{lst:unknown_phylogeny}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing \ref{lst:create_alignment}.

@@

The resulting figure is shown in figure \ref{fig:example_5}.

@@

However if the same procedure were repeated and performed on a distribution 
with a sufficient number of generative trees, 
it could actually constitute a quantitative assessment 
of the quality of the inference.
To give an idea of the insights \verb;pirouette; can give,
we pretend the results of the example research questions
to be obtained from thousands of replicates.
In that case, figure \ref{fig:example_1} shows 
the minimum error BEAST2 when the
generative model is known, which can serve as 
a minimum noise level assessment in a bigger experiment.  
Figure \ref{fig:example_2} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the best candidate model gives an error distribution similar to the unknown tree prior,
hinting that the candidate model would be just as good in inference.
Figure \ref{fig:example_3} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the twin model gives an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2.
Figure \ref{fig:example_4} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the twin model and best candidate model give 
an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2, 
and that the candidate model would be just as good in inference.

>>>>

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree
prior. It does so by measuring the inference error BEAST2
makes for a given reconstructed phylogeny, 
simulated under the mechanisms of a (usually novel) diversification model, 
which we will refer to as the 'generative tree prior', $\mathit{P}$. 
Such a study will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the error made by BEAST2 cannot be neglected. 
Although a well-conducted theoretical study uses many replicates, 
this article only shows examples with one replicate, for the sake of clarity.

\verb;pirouette; is very flexible and allows the user 
to specify a big variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in tables~\ref{tab:options} and~\ref{tab:definitions}.

@@

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
We will exploit these functions to answer our research questions in the next sections.

@@

\begin{lstlisting}[
    language=R, floatplacement=ht, frame=single, 
    label = {lst:create_yule_tree}, 
    caption = {
      Create a Yule tree. 
      The resulting tree is shown in figure~\ref{fig:yule_tree}
    }
  ]
phylogeny <- create_yule_tree(n_taxa = 6, crown_age = 10)
\end{lstlisting}

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_yule.png}
  \caption{The Yule tree, as created by listing~\ref{lst:create_yule_tree}.}
  \label{fig:yule_tree}
  \giovanni{Remove the text in top-left corner.}
  \giovanni{I'd use a sexier topology, purely for esthetical reasons.}
\end{figure}

@@

The resulting figure is shown in figure~\ref{fig:example_1}.

@@

\begin{lstlisting}[
  language=R, 
  floatplacement=ht, 
  frame=single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[ht]
  \includegraphics[width=\textwidth]{tree_unknown.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
  }
\end{figure}

The first step in \verb;pirouette; is to simulate a DNA alignment from the given phylogeny. 
We will re-use the alignment parameters of the previous example 
as shown in listing~\ref{lst:create_alignment}.

@@

The resulting figure is shown in figure~\ref{fig:example_2}.

@@

Most of the settings for this example research question are
the same as in the previous section, which are
the phylogeny from an unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
the alignment parameters (listing~\ref{lst:create_alignment}) 
and the list of experiments (listing~\ref{lst:create_generative_experiment}).

@@

The resulting figure is shown in figure~\ref{fig:example_3}

@@

Here, same as example 2 and 3, we start with the phylogeny 
from an unknown speciation model (listing~\ref{lst:unknown_phylogeny}), 
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

In the second step we specify we want all 
40 inference models to compete, by 
calling \verb;create_all_experiments;,
as done in listing~\ref{lst:create_all_experiments}.

@@

To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using the
default created twinning parameters, as shown in 
listing~\ref{lst:create_twinning_params}.

@@

The resulting figure is shown in figure~\ref{fig:example_4}.

@@

In this example, we use a tree generated by the Yule process.
Instead of using the common Yule tree as in example 1, we'll use
a Yule tree of low (actually: zero) likelihood: 
the one as used in examples 2-4.
by using the code shown in listing~\ref{lst:unknown_phylogeny}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

The resulting figure is shown in figure~\ref{fig:example_5}.

@@

However if the same procedure were repeated and performed on a distribution 
with a sufficient number of generative trees, 
it could actually constitute a quantitative assessment 
of the quality of the inference.
To give an idea of the insights \verb;pirouette; can give,
we pretend the results of the example research questions
to be obtained from thousands of replicates.
In that case, figure~\ref{fig:example_1} shows 
the minimum error BEAST2 when the
generative model is known, which can serve as 
a minimum noise level assessment in a bigger experiment.  
Figure~\ref{fig:example_2} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the best candidate model gives an error distribution similar to the unknown tree prior,
hinting that the candidate model would be just as good in inference.
Figure~\ref{fig:example_3} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the twin model gives an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2.
Figure~\ref{fig:example_4} shows that 
\richel{
  reword after new results with 6 taxa tree
} 
the twin model and best candidate model give 
an error distribution similar to the unknown tree prior,
hinting that the shape of the phylogeny with unknown tree prior
does not effect the error made by BEAST2, 
and that the candidate model would be just as good in inference.
